---
title: "HEV Luminex QC Report"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: false
params:
  plate_file: null
  value_datatype: "Net MFI"
  cv_threshold: 0.20
  blank_quantile: 0.99
  min_total_events: 50
  bead_threshold: 30
  history_rep_path: "outputs/history/replicate_history.rds"
  history_blank_path: "outputs/history/blank_history.rds"
  std_history_path: "outputs/history/std_curve_history.rds"
  control_history_path: "outputs/history/control_mfi_history.rds"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,        # hide R code
  message = FALSE,     # hide package attach messages
  warning = FALSE,     # hide warnings like "built under R version"
  error = TRUE,        # show errors if they happen
  fig.align = "center"
)

# load packages
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(gt)
library(DT)

# Project functions
source("R/read_xponent_plate.R")
source("R/qc_replicates.R")
source("R/qc_plate_summary.R")
source("R/qc_history.R")
source("R/qc_standards.R")
source("R/qc_beads.R")
source("R/qc_standards_history.R")
source("R/standard_4pl.R")
source("R/subtract_blanks.R")


# Order plate_id factor levels by the trailing plate number
# e.g. PLATE_01292026_plate2 -> 2, PLATE_02192026_plate17 -> 17
order_plate_id <- function(x) {
  plate_num <- as.numeric(str_extract(x, "(?<=plate)\\d+$"))
  lvls <- unique(x[order(plate_num, na.last = TRUE)])
  factor(x, levels = lvls)
}

gt_theme_qc <- function(gt_tbl) {
  gt_tbl %>%
    opt_table_outline() %>%
    opt_vertical_padding(scale = 0.75) %>%
    tab_options(
      table.font.size = px(13),
      heading.title.font.size = px(16),
      column_labels.font.weight = "bold",
      data_row.padding = px(4)
    )
}

# helper function (should probably add to a sperate R file)
extract_control_points <- function(df) {
  ctrl <- df %>%
    mutate(
      well_type = case_when(
        str_detect(str_to_lower(sample), "\\bpos\\b") ~ "control_pos",
        str_detect(str_to_lower(sample), "\\bneg\\b") ~ "control_neg",
        TRUE ~ NA_character_
      )
    ) %>%
    filter(!is.na(well_type), !is.na(value))

  # compute 1/RAU if norm_denom is available (i.e. after 4PL normalization)
  if ("norm_denom" %in% names(ctrl)) {
    ctrl <- ctrl %>%
      mutate(
        dil_denom = case_when(
          str_detect(dilution, "^1/") ~ as.numeric(sub("^1/", "", dilution)),
          str_to_lower(dilution) %in% c("neat", "undiluted", "1") ~ 1,
          TRUE ~ NA_real_
        ),
        inv_rau = if_else(
          !is.na(norm_denom) & norm_denom > 0 & !is.na(dil_denom),
          norm_denom / dil_denom,
          NA_real_
        )
      )
  }

  result <- ctrl %>%
    group_by(plate_id, run_date, analyte, dilution, well_type) %>%
    summarise(
      mfi = median(value, na.rm = TRUE),
      n = dplyr::n(),
      .groups = "drop"
    )

  if ("inv_rau" %in% names(ctrl)) {
    rau_summary <- ctrl %>%
      group_by(plate_id, run_date, analyte, dilution, well_type) %>%
      summarise(
        median_inv_rau = median(inv_rau, na.rm = TRUE),
        .groups = "drop"
      )
    result <- left_join(result, rau_summary,
                        by = c("plate_id", "run_date", "analyte", "dilution", "well_type"))
  }

  result
}
```


```{r}
stopifnot(!is.null(params$plate_file))


df <- read_xponent_plate(params$plate_file, value_datatype = params$value_datatype)
#df <- read_xponent_plate("data/raw/PlateRunResults_PLATE_01292026_plate2.csv", value_datatype = params$value_datatype)



# Safety: drop placeholder analytes
df <- df %>%
  filter(!str_detect(analyte, "^X(_\\d+)?$"), analyte != "", !is.na(analyte))

# Compute plate summary BEFORE blank subtraction so that blank QC stats
# (q99, median) reflect the raw signal levels for display and history tracking
ps <- plate_summary(
  df,
  min_total_events = params$min_total_events,
  blank_quantile = params$blank_quantile
)

# Subtract mean blank per plate/analyte from all values to get true Net MFI.
# xPONENT's "Net MFI" is actually Median because blanks were not identified
# in the instrument. Negative values are floored at 0.
df <- subtract_plate_blanks(df)

rep_tbl <- compute_replicate_variability(
  df,
  cv_threshold = params$cv_threshold,
  blank_quantile = params$blank_quantile,
  low_signal_rule = "blank_q99"
)

# History
rep_hist_old   <- history_load(params$history_rep_path)
blank_hist_old <- history_load(params$history_blank_path)
hist_records <- make_history_records(df, rep_tbl, ps$blanks)
rep_hist_new <- history_append(rep_hist_old, hist_records$rep_hist,
                               key_cols = c("plate_id","analyte"))
blank_hist_new <- history_append(blank_hist_old, hist_records$blank_hist,
                                 key_cols = c("plate_id","analyte"))

history_save(rep_hist_new, params$history_rep_path)
history_save(blank_hist_new, params$history_blank_path)

## standards
std_hist_path <- params$std_history_path
std_hist_old <- std_history_load(std_hist_path)

std_pts <- extract_std_points(df)
std_hist_new <- std_history_append(std_hist_old, std_pts)

std_history_save(std_hist_new, std_hist_path)


current_plate_id <- df %>% summarise(pid = first(plate_id)) %>% pull(pid)

plate_meta <- df %>%
  summarise(
    plate_id = first(plate_id),
    plate_batch = first(plate_batch),
    run_date = first(run_date),
    panel = first(panel),
    file = first(plate_file),
    datatype = first(value_datatype)
  )

```

## Plate overview

```{r}
plate_meta %>%
  pivot_longer(everything(), names_to = "Field", values_to = "Value") %>%
  gt() %>%
  tab_header(title = "Plate overview") %>%
  cols_width(Field ~ px(200), Value ~ px(520)) %>%
  gt_theme_qc()
```

### Sample counts

The table below shows the number of unique samples of each sample type and dilution (when applicable) on the plate.
```{r}
ps$counts %>%
  arrange(well_type) %>%
  gt() %>%
  tab_header(title = "Sample counts") %>%
  cols_label(well_type = "Sample type", n = "N") %>%
  gt_theme_qc()
```

### Bead Summary

How many beads of each type are present on the plate?

```{r}
expected_beads <- 96

ps$analytes %>%
  gt() %>%
  tab_header(title = "Bead summary") %>%
  cols_label(analyte = "Bead type", n = "N") %>%
tab_style(
  style = cell_text(color = "red", weight = "bold"),
  locations = cells_body(
    columns = n,
    rows = n < expected_beads
  )
) %>%
gt_theme_qc()
```


## Bead counts

```{r}
bead_flags <- df %>%
  filter(!is.na(bead_count)) %>%
  mutate(low_beads = bead_count < 30)

bead_summary <- bead_flags %>%
  summarise(
    n_points = n(),
    n_low = sum(low_beads),
    pct_low = mean(low_beads),
    min_beads = min(bead_count, na.rm = TRUE)
  )

bead_summary  %>%
  gt()
```

### Wells/analytes with bead count < 30

```{r}
bead_flags %>%
  filter(low_beads) %>%
  distinct(well, sample, dilution, well_type, analyte, bead_count) %>%
  arrange(bead_count) %>%
  gt()
```

### Distribution of bead counts

```{r}
ggplot(bead_flags, aes(x = bead_count)) +
  geom_histogram(bins = 50) +
  labs(x = "Bead count (Count DataType)", y = "Number of well×analyte points")
```



## Replicate variability

### Summary

```{r}
rep_summary <- rep_tbl %>%
  summarise(
    n_pairs = n(),
    pct_fail = mean(flag == "fail_high_cv"),
    pct_missing = mean(flag == "warn_missing_rep")
  )

rep_summary %>%
  gt() %>%
  tab_header(title = "Replicate variability summary") %>%
  cols_label(
    n_pairs = "Replicate pairs",
    pct_fail = "% failing CV",
    pct_missing = "% missing replicates"
  ) %>%
  fmt_percent(c(pct_fail, pct_missing), decimals = 1) %>%
  gt_theme_qc()
```

### Worst replicate pairs

```{r}
rep_tbl %>%
  arrange(desc(cv_primary)) %>%
  select(sample, dilution, analyte, rep1, rep2, mean_raw, cv_primary, flag) %>%
  head(30) %>%
  gt() %>%
  tab_header(title = "Worst replicate pairs (top 30)") %>%
  fmt_number(c(rep1, rep2, mean_raw, cv_primary), decimals = 3) %>%
  tab_style(
    style = cell_text(weight = "bold", color = "firebrick"),
    locations = cells_body(columns = flag, rows = flag != "pass")
  ) %>%
  gt_theme_qc()
```

### CV distribution by analyte

```{r}
ggplot(rep_tbl, aes(x = cv_primary)) +
  geom_histogram(bins = 40) +
  facet_wrap(~ analyte, scales = "free_y") +
  labs(x = "Primary CV (raw unless low-signal, then log-CV)", y = "Count")
```

### Replicate scatter (log scale)

```{r}
ggplot(rep_tbl, aes(x = rep1, y = rep2)) +
  geom_point(alpha = 0.6) +
  scale_x_log10() +
  scale_y_log10() +
  geom_abline(intercept = 0, slope = 1, 
            linetype = "dashed", color = "grey50") +
  facet_wrap(~ analyte) +
  labs(x = "Replicate 1 (Net MFI, blank-subtracted)", y = "Replicate 2 (Net MFI, blank-subtracted)")
```

## Blanks

```{r}
ps$blanks %>%
  arrange(analyte) %>%
  gt() %>%
  tab_header(title = "Blank signal by analyte") %>%
  fmt_number(c(blank_q99, blank_median), decimals = 2) %>%
  gt_theme_qc()
```

```{r}
ggplot(ps$blanks, aes(x = analyte, y = blank_q99)) +
  geom_point() +
  coord_flip() +
  labs(x = "Analyte", y = "Blank 99th percentile (raw MFI, pre-subtraction)")
```

## Across-plate trends (history)

### Replicate fail rate over time

```{r}
rep_hist_new %>%
  mutate(run_date = as.character(run_date)) %>%
  ggplot(aes(x = run_date, y = pct_fail, group = analyte)) +
  geom_point() +
  facet_wrap(~ analyte, scales = "free_y") +
  labs(x = "Run date", y = "Pct replicate pairs failing (CV >= threshold)")
```

### Blank q99 trend over time

```{r}
blank_hist_new %>%
  mutate(plate_id = order_plate_id(plate_id)) %>%
  ggplot(aes(x = plate_id, y = value, group = analyte)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ analyte, scales = "free_y") +
  labs(x = "Plate ID", y = "Blank q99 (raw MFI, pre-subtraction)") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

## Outputs (tables)

```{r}
out_dir <- "outputs/tables"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

rep_out <- file.path(out_dir, paste0("replicate_metrics_", plate_meta$plate_id, ".csv"))
write.csv(rep_tbl, rep_out, row.names = FALSE)

flags <- rep_tbl %>%
  filter(flag != "pass") %>%
  transmute(
    plate_id, sample, dilution, analyte,
    issue = flag,
    cv_primary
  )

flags_out <- file.path(out_dir, paste0("qc_flags_", plate_meta$plate_id, ".csv"))
write.csv(flags, flags_out, row.names = FALSE)

list(replicate_metrics = rep_out, qc_flags = flags_out)
```


## Standard curves (per plate)
```{r}
std_df <- extract_standards(df)
normalize_analytes <- c(
  "ORF2 Bohm", "ORF2 Wantai", "ORF2-E2 Bohm",
  "ORF2ID", "ORF3 Bohm"
)

fits_4pl <- fit_4pl_all(std_df)
fits_4pl <- fits_4pl %>%
  mutate(
    use_for_norm = fit_ok & analyte %in% normalize_analytes
  )

# append normalized values to the main dataset
df <- normalize_df_to_4pl(df, fits_4pl)

# Controls history (after normalization so norm_denom / RAU is available)
ctrl_hist_path <- params$control_history_path
ctrl_hist_old <- history_load(ctrl_hist_path)

ctrl_pts <- extract_control_points(df)
ctrl_hist_new <- history_append(
  ctrl_hist_old, ctrl_pts,
  key_cols = c("plate_id", "analyte", "dilution", "well_type")
)

history_save(ctrl_hist_new, ctrl_hist_path)

df %>%
  summarise(
    n_specimens = sum(sample_type == "specimen"),
    n_specimens_with_norm = sum(sample_type == "specimen" & !is.na(norm_denom)),
    pct_with_norm = n_specimens_with_norm / n_specimens
  )

std_levels <- std_df %>%
  filter(!is_blank, !is.na(denom)) %>%
  distinct(denom) %>%
  arrange(denom)
```

### Standard curve fits (just to make sure)
```{r}
fits_4pl %>%
  mutate(note = if_else(fit_ok, "", "fit failed / insufficient points")) %>%
  select(plate_id, analyte, fit_ok, a, b, c, d, note)
```

### Standard curves by analyte
```{r fig.width=12, fig.height=7}

# ---- current plate standards (points) ----
std_current <- std_df %>%
  filter(
    plate_id == unique(df$plate_id),
    !is_blank,
    !is.na(denom),
    denom > 0
  )

# ---- current plate specimens ----
spec_current <- df %>%
  filter(plate_id == current_plate_id, well_type == "specimen", !is.na(value)) %>%
  select(analyte, value, dilution)

# ---- historical observed standards (grey lines) ----
std_hist <- readRDS("outputs/history/std_curve_history.rds") %>%
  filter(
    plate_id != unique(df$plate_id),
    !is.na(denom),
    denom > 0
  )

# Optional: summarise replicates before connecting
std_hist_summary <- std_hist %>%
  group_by(plate_id, analyte, denom) %>%
  summarise(
    value = median(value, na.rm = TRUE),
    .groups = "drop"
  )

# ---- fitted curve for current plate ----
curve_current <- fits_4pl %>%
  filter(fit_ok) %>%
  rowwise() %>%
  mutate(
    denom = list(exp(seq(
      log(min(std_current$denom[std_current$analyte == analyte])),
      log(max(std_current$denom[std_current$analyte == analyte])),
      length.out = 200
    )))
  ) %>%
  tidyr::unnest(denom) %>%
  mutate(
    fitted = d + (a - d) / (1 + (denom / c)^b)
  )

# ---- plot ----
p <- ggplot() +
  # historical observed curves (grey)
  geom_line(
    data = std_hist_summary,
    aes(
      x = denom,
      y = value,
      group = plate_id
    ),
    color = "grey80",
    linewidth = 0.5,
    alpha = 0.7
  ) +
  # current fitted curve (black)
  geom_line(
    data = curve_current,
    aes(x = denom, y = fitted),
    color = "black",
    linewidth = 1.2
  ) +
  # current plate points (replicates separate)
  geom_point(
    data = std_current,
    aes(x = denom, y = value, shape = factor(rep_index)),
    size = 2.2,
    alpha = 0.9
  ) +
  scale_x_log10() +
  facet_wrap(~ analyte, scales = "free_y") +
  labs(
    x = "Standard dilution denominator (log scale)",
    y = "Net MFI (blank-subtracted)",
    shape = "Replicate",
    title = "Standard curves: observed historical data (grey) and current plate fit (black)"
  ) +
  theme(legend.position = "bottom") + 
  theme_minimal() + 
  geom_rug(
    data = spec_current,
    aes(y = value, color = dilution),
    sides = "r",
    alpha = 0.35,
    show.legend = TRUE
  ) +
  guides(color = guide_legend(title = "Specimen dilution"))
 
print(p)
```

## Controls by plate

```{r}
pdat <- ctrl_hist_new %>%
  mutate(
    is_current = plate_id == current_plate_id,
    plate_id = order_plate_id(plate_id)
  )

pdat %>%
  ggplot(aes(x = plate_id, y = mfi, group = interaction(well_type, dilution))) +
  # historical
  geom_point(data = ~dplyr::filter(.x, !is_current),
             aes(color = well_type, shape = dilution, alpha = 0.6)) +
  # current highlighted
  geom_point(data = ~dplyr::filter(.x, is_current),
             aes(color = well_type, shape = dilution)) +
  facet_wrap(~ analyte, scales = "free_y") +
  scale_y_log10() +
  labs(x = "Plate", y = "Control Net MFI (median, log scale, blank-subtracted)",
       color = "Control", shape = "Dilution") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

### Positive control MFI

```{r}
ctrl_pos_hist <- ctrl_hist_new %>%
  filter(well_type == "control_pos") %>%
  mutate(
    is_current = plate_id == current_plate_id,
    plate_id = order_plate_id(plate_id)
  )
```

```{r}
ctrl_pos_hist %>%
  select(plate_id, run_date, analyte, dilution, mfi) %>%
  arrange(plate_id, analyte, dilution) %>%
  mutate(mfi = round(mfi, 1)) %>%
  datatable(
    caption = "Positive control MFI by plate, analyte, and dilution",
    extensions = "Buttons",
    options = list(
      pageLength = 10,
      dom = "Bfrtip",
      buttons = c("csv", "excel")
    ),
    rownames = FALSE
  )
```

```{r fig.width=12, fig.height=7}
ggplot(ctrl_pos_hist, aes(x = plate_id, y = mfi, color = dilution)) +
  geom_point(aes(alpha = is_current), size = 2.5) +
  scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.4), guide = "none") +
  facet_wrap(~ analyte, scales = "free_y") +
  scale_y_log10() +
  labs(
    x = "Plate", y = "Positive control Net MFI (median, log scale, blank-subtracted)",
    color = "Dilution",
    title = "Positive control MFI across plates"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

### Positive control 1/RAU

```{r}
ctrl_pos_rau_hist <- ctrl_hist_new %>%
  filter(well_type == "control_pos", !is.na(median_inv_rau)) %>%
  mutate(
    is_current = plate_id == current_plate_id,
    plate_id = order_plate_id(plate_id)
  )
```

```{r}
ctrl_pos_rau_hist %>%
  select(
    Plate = plate_id,
    `Run date` = run_date,
    Analyte = analyte,
    Dilution = dilution,
    `Median 1/RAU` = median_inv_rau
  ) %>%
  arrange(Plate, Analyte, Dilution) %>%
  mutate(`Median 1/RAU` = round(`Median 1/RAU`, 2)) %>%
  datatable(
    caption = "Positive control 1/RAU by plate, analyte, and dilution",
    extensions = "Buttons",
    options = list(
      pageLength = 10,
      dom = "Bfrtip",
      buttons = c("csv", "excel")
    ),
    rownames = FALSE
  )
```

```{r fig.width=12, fig.height=7}
ggplot(ctrl_pos_rau_hist, aes(x = plate_id, y = median_inv_rau, color = dilution)) +
  geom_point(aes(alpha = is_current), size = 2.5) +
  scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.4), guide = "none") +
  facet_wrap(~ analyte, scales = "free_y") +
  scale_y_log10() +
  labs(
    x = "Plate", y = "Positive control 1/RAU (median, log scale)",
    color = "Dilution",
    title = "Positive control 1/RAU across plates"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

## Specimen MFIs and RAUs

```{r fig.width=12, fig.height=7}
plot_df <- df %>%
  filter(
    well_type == "specimen",
    !is.na(value),
    value > 0,
    !is.na(dilution)
  )

p <- ggplot(
  plot_df,
  aes(x = value, color = dilution, fill = dilution)
) +
  geom_histogram(
    bins = 50,
    alpha = 0.35,
    position = "identity"
  ) +
  scale_x_log10() +
  facet_wrap(~ analyte, scales = "free_y") +
  labs(
    x = "Net MFI (log scale, blank-subtracted)",
    y = "Count",
    color = "Dilution",
    fill = "Dilution",
    title = "Specimen Net MFI distribution by antigen and dilution (blank-subtracted)"
  ) +
  theme(legend.position = "bottom")

print(p)
```

```{r fig.width=12, fig.height=7}
plot_df <- df %>%
  filter(
    sample_type == "specimen",
    !is.na(dilution),
    !is.na(norm_denom),
    norm_denom > 0
  ) %>%
  mutate(
    dil_denom = case_when(
      str_detect(dilution, "^1/") ~ as.numeric(sub("^1/", "", dilution)),
      str_to_lower(dilution) %in% c("neat", "undiluted", "1") ~ 1,
      TRUE ~ NA_real_
    ),
    rau = dil_denom / norm_denom,
    inv_rau = 1 / rau   # == norm_denom / dil_denom
  )

ggplot(plot_df, aes(x = inv_rau, color = dilution, fill = dilution)) +
  geom_histogram(bins = 50, alpha = 0.35, position = "identity") +
  scale_x_log10() +
  facet_wrap(~ analyte, scales = "free_y") +
  labs(
    x = "1 / RAU (dilution-adjusted, log scale)",
    y = "Count",
    color = "Dilution",
    fill = "Dilution",
    title = "Specimen 1/RAU (dilution-adjusted) by antigen and dilution"
  ) +
  theme(legend.position = "bottom")
```
## To dos:
- Check on LL4 fits and predictions
- Redo RAU plots of analytes